---
title: "Data Wrangling-Air Quality"
author: "Dwiti Bagadia and Yue Zhang"
date: "2023-04-18"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed
#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=40), tidy=TRUE)

```

## OVERVIEW

This R Markdown file aims to wrangle the data of New York State air quality by county from 2018 to 2020.


## Set working directory and load packages
```{r packages, message = FALSE}
setwd("E:/MCRP/Spring2023/BagadiaZhang_ENV872_EDA_FinalProject")
getwd()
library(tidyverse)
library(lubridate)
library(dplyr)
```

## Load AQ data and set working directory
```{r data, message = FALSE}
US_AQ_2018 = read.csv("./Data/RawData/annual_aqi_by_county_2018.csv")
US_AQ_2019 = read.csv("./Data/RawData/annual_aqi_by_county_2019.csv")
US_AQ_2020 = read.csv("./Data/RawData/annual_aqi_by_county_2020.csv")
US_Ozone = read.csv("./Data/RawData/daily_44201_2020.csv")
US_SO2 = read.csv("./Data/RawData/daily_42401_2020.csv")
US_CO = read.csv("./Data/RawData/daily_42101_2020.csv")
US_NO2 = read.csv("./Data/RawData/daily_42602_2020.csv")
US_PM25 = read.csv("./Data/RawData/daily_88502_2020.csv")
US_PM10 = read.csv("./Data/RawData/daily_81102_2020.csv")

```

## Select NY AQ data
```{r NY data, message = FALSE}
NY_AQ_2018 = subset(US_AQ_2018, State == "New York")
NY_AQ_2019 = subset(US_AQ_2019, State == "New York")
NY_AQ_2020 = subset(US_AQ_2020, State == "New York")

```

## Clean NY data and calculate days with bad air quality
```{r clean NY data, message = FALSE}
NY_AQ_2018_clean = NY_AQ_2018 %>% 
  mutate("BadDays" = Days.with.AQI - Good.Days) %>% 
  select(County, Days.with.AQI, BadDays)
colnames(NY_AQ_2018_clean) = c("County", "Days18", "BadDays18")
NY_AQ_2019_clean = NY_AQ_2019 %>% 
  mutate("BadDays" = Days.with.AQI - Good.Days) %>% 
  select(County, Days.with.AQI, BadDays)
colnames(NY_AQ_2019_clean) = c("County", "Days19", "BadDays19")
NY_AQ_2020_clean = NY_AQ_2020 %>% 
  mutate("BadDays" = Days.with.AQI - Good.Days) %>% 
  select(County, Days.with.AQI, BadDays)
colnames(NY_AQ_2020_clean) = c("County", "Days20", "BadDays20")

```

## Join the NY AQ datasets
```{r Join, message = FALSE}
NY_AQ_join = merge(NY_AQ_2018_clean, NY_AQ_2019_clean, by = "County") %>% 
  merge(NY_AQ_2020_clean, by = "County")

```

## Calculate the mean bad days from 2018 to 2020
```{r clean NYC data, message = FALSE}
NY_AQ_join_clean = NY_AQ_join %>% mutate("BadDaysRate" = as.numeric(BadDays18 + BadDays19 + BadDays20)*100/as.numeric(Days18 + Days19 + Days20)) %>% select(County, BadDaysRate)

```

## Select NY Ozone data
```{r NY data, message = FALSE}
NY_Ozone = subset(US_Ozone, State.Name == "New York")

```

## Clean NY data and calculate mean ozone AQI
```{r clean NY data, message = FALSE}
# AQI in this file means the Air Quality Index for the day for ozone
NY_Ozone_clean = NY_Ozone[, c(24, 6:8, 19)]
NY_Ozone_clean = NY_Ozone_clean %>% 
  group_by(County.Name) %>% 
  summarise(meanOzone = mean(AQI))
colnames(NY_Ozone_clean) = c("County", "meanOzone")

```

## Select NY SO2 data
```{r NY data, message = FALSE}
NY_SO2 = subset(US_SO2, State.Name == "New York")

```

## Clean NY data and calculate mean SO2 AQI
```{r clean NY data, message = FALSE}
# AQI in this file means the Air Quality Index for the day for SO2
NY_SO2_clean = NY_SO2[, c(26, 6:8, 10, 20)]
NY_SO2_clean = NY_SO2_clean %>% 
  filter(Sample.Duration == "1 HOUR") %>% 
  group_by(County.Name) %>% 
  summarise(meanSO2 = mean(AQI))
colnames(NY_SO2_clean) = c("County", "meanSO2")

```

## Select NY CO data
```{r NY data, message = FALSE}
NY_CO = subset(US_CO, State.Name == "New York")

```

## Clean NY data and calculate mean CO AQI
```{r clean NY data, message = FALSE}
# AQI in this file means the Air Quality Index for the day for CO
NY_CO_clean = NY_CO[, c(26, 6:8, 10, 20)]
NY_CO_clean = NY_CO_clean %>% 
  filter(Sample.Duration == "8-HR RUN AVG END HOUR") %>% 
  group_by(County.Name) %>% 
  summarise(meanCO = mean(AQI))
colnames(NY_CO_clean) = c("County", "meanCO")

```

## Select NY NO2 data
```{r NY data, message = FALSE}
NY_NO2 = subset(US_NO2, State.Name == "New York")

```

## Clean NY data and calculate mean NO2 AQI
```{r clean NY data, message = FALSE}
# AQI in this file means the Air Quality Index for the day for NO2
NY_NO2_clean = NY_NO2[, c(26, 6:8, 20)]
NY_NO2_clean = NY_NO2_clean %>% 
  group_by(County.Name) %>% 
  summarise(meanNO2 = mean(AQI))
colnames(NY_NO2_clean) = c("County", "meanNO2")

```

## Select NY PM2.5 data
```{r NY data, message = FALSE}
NY_PM25 = subset(US_PM25, State.Name == "New York")

```

## Clean NY data and calculate mean PM2.5 AQI
```{r clean NY data, message = FALSE}
# AQI in this file means the Air Quality Index for the day for PM2.5
NY_PM25_clean = NY_PM25[, c(26, 6:8, 10, 20)]
NY_PM25_clean = NY_PM25_clean %>% 
  filter(Sample.Duration == "24-HR BLK AVG") %>% 
  group_by(County.Name) %>% 
  summarise(meanPM25 = mean(AQI))
colnames(NY_PM25_clean) = c("County", "meanPM25")

```

## Select NY PM10 data
```{r NY data, message = FALSE}
NY_PM10 = subset(US_PM10, State.Name == "New York")

```

## Clean NY data and calculate mean PM10 AQI
```{r clean NY data, message = FALSE}
# AQI in this file means the Air Quality Index for the day for PM10 
NY_PM10_clean = NY_PM10[, c(26, 6:8, 20)]
NY_PM10_clean = NY_PM10_clean %>% 
  group_by(County.Name) %>% 
  summarise(meanPM10 = mean(AQI))
colnames(NY_PM10_clean) = c("County", "meanPM10")

```

## Save NY data
```{r save data, message = FALSE}
write.csv(NY_AQ_join_clean, "./Data/ProcessedData/NY_AQ_processed1.csv", row.names = FALSE)
write.csv(NY_Ozone_clean, "./Data/ProcessedData/NY_Ozone_processed.csv", row.names = FALSE)
write.csv(NY_SO2_clean, "./Data/ProcessedData/NY_SO2_processed.csv", row.names = FALSE)
write.csv(NY_CO_clean, "./Data/ProcessedData/NY_CO_processed.csv", row.names = FALSE)
write.csv(NY_NO2_clean, "./Data/ProcessedData/NY_NO2_processed.csv", row.names = FALSE)
write.csv(NY_PM25_clean, "./Data/ProcessedData/NY_PM25_processed.csv", row.names = FALSE)
write.csv(NY_PM10_clean, "./Data/ProcessedData/NY_PM10_processed.csv", row.names = FALSE)

```
