---
title: "Linear Regression for Asthma and Air Quality"
author: "Dwiti Bagadia and Yue Zhang"
date: "2023-04-24"
output: pdf_document
---

```{r setup, include = FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed
#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=40), tidy=TRUE)

```

## OVERVIEW

This R Markdown file aims to generate the linear model for asthma rate and air quality in New York State in 2020.

## Set working directory and load packages
```{r packages, message = FALSE}
setwd("E:/MCRP/Spring2023/BagadiaZhang_ENV872_EDA_FinalProject")
getwd()
library(tidyverse)
library(lubridate)
library(dplyr)
library(agricolae)
library(ggthemes)

```

## Load data and set plot themes
```{r data, message = FALSE}
NY_Asthma_processed = read.csv("./Data/ProcessedData/NY_Asthma_processed1.csv")
NY_AQ_processed = read.csv("./Data/ProcessedData/NY_AQ_processed1.csv")
mytheme = theme_clean(base_size = 12) + 
  theme(axis.text = element_text(color = "black"), legend.position = "right", axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5), plot.title = element_text(size = 12))
theme_set(mytheme)
```

## Join the asthma data and air quality data together
```{r join, message = FALSE}
NY_Asthma_AQ_join = left_join(NY_AQ_processed, NY_Asthma_processed, by = "County")

```

## Visualize the relationship of bad days rate by asthma rate with a scatter plot
```{r scatter plot, message = FALSE}
AsthmaAQ.points = ggplot(NY_Asthma_AQ_join, aes(y = log(BadDaysRate), x =VisitRate)) + 
  geom_point(aes(color = County)) + 
  geom_smooth(method = "lm") + 
  labs(y = "Bad Days Rate", x = "Emergency Visit Rate (per 10k)")
print(AsthmaAQ.points)

```

## Generate linear model to test the relationship
```{r linear.regression, message = FALSE}
AsthmaAQ.regression = lm(data = NY_Asthma_AQ_join, log(BadDaysRate) ~ VisitRate)
summary(AsthmaAQ.regression)
plot(AsthmaAQ.regression,1)
plot(AsthmaAQ.regression,2)
plot(AsthmaAQ.regression,3)
plot(AsthmaAQ.regression,4)

```