---
title: "Spatial Analysis"
author: "Dwiti Bagadia and Yue Zhang"
date: "2023-05-02"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Overview
This R markdown file aims to create maps and show spatial distribution of air quality and asthma rates.

## Load packages and set working directory
```{r load packages, message = FALSE}
library(tidyverse)
library(sf)
library(leaflet)
library(mapview)
library(RColorBrewer)
setwd("E:/MCRP/Spring2023/BagadiaZhang_ENV872_EDA_FinalProject")
getwd()

```

## Load data
```{r load data, message = FALSE}
NY_Asthma_processed = read.csv("./Data/ProcessedData/NY_Asthma_processed1.csv")
NY_AQ_processed = read.csv("./Data/ProcessedData/NY_AQ_processed1.csv")
NY_Demographic_processed = read.csv("./Data/ProcessedData/NY_Demographic_processed1.csv")
NY_Ozone_processed = read.csv("./Data/ProcessedData/NY_Ozone_processed.csv")
NY_PM25_processed = read.csv("./Data/ProcessedData/NY_PM25_processed.csv")
NY_SO2_processed = read.csv("./Data/ProcessedData/NY_SO2_processed.csv")
```

## Read the counties shapefile and filter NY and visualize it 
```{r data exploratory, message = FALSE}
NY_sf = st_read("./Data/SpatialData/cb_2018_us_county_20m.shp") %>% filter(STATEFP == "36")
st_crs(NY_sf)
head(NY_sf)
NY_sf_2263 = st_transform(NY_sf, crs = 2263)
mapView(NY_sf_2263)

```

## Join asthma data to counties shapefile and visualize it 
```{r asthma visualize, message = FALSE}
NY_Asthma_Counties_sf = merge(NY_sf_2263,
                              NY_Asthma_processed,
                              by.x = "NAME",
                              by.y = "County")
mapView(
  NY_Asthma_Counties_sf,
  zcol = "VisitRate",
  col.regions = brewer.pal(10, "RdBu"),
  layer.name = "Asthma Rate per 10,000 Children in New York Stae from 2018 to 2020",
  map.types = "CartoDB.Positron",
  label = NY_Asthma_Counties_sf$NAME,
  labelOptions = labelOptions(
    noHide = TRUE,
    textOnly = TRUE,
    textsize = "8px",
    direction = "center"
  )
)

```

## Join demographic data to counties shapefile and visualize it 
```{r demographic visualize, message = FALSE}
NY_Demographic_Counties_sf = merge(NY_sf_2263,
                                   NY_Demographic_processed,
                                   by.x = "NAME",
                                   by.y = "County")
mapView(
  NY_Demographic_Counties_sf,
  zcol = "NonWhiteDens",
  col.regions = brewer.pal(10, "PiYG"),
  layer.name = "Non-White Population Density per Sq.mile in New York State from 2016 to 2020",
  map.types = "CartoDB.Positron",
  label = NY_Demographic_Counties_sf$NAME,
  labelOptions = labelOptions(
    noHide = TRUE,
    textOnly = TRUE,
    textsize = "8px",
    direction = "center"
  )
) 

```

## Join pollutants data to counties shapefile and visualize it 
```{r pollutants visualize, message = FALSE}
#Ozone monitor sites
NY_Ozone_Counties_sf = merge(NY_sf_2263, NY_Ozone_processed, by.x = "NAME", by.y = "County")
NY_Ozone_point = st_centroid(NY_Ozone_Counties_sf)
st_crs(NY_Ozone_point)
mapView(NY_sf_2263, col.regions = "orange") + mapView(
  NY_Ozone_point,
  cex = "meanOzone",
  col.regions = "mediumpurple",
  layer.name = "Ozone in New York State in 2020",
  map.types = "CartoDB.Positron"
)

#PM2.5 monitor sites
NY_PM25_Counties_sf = merge(NY_sf_2263, NY_PM25_processed, by.x = "NAME", by.y = "County")
NY_PM25_point = st_centroid(NY_PM25_Counties_sf)
mapView(NY_sf_2263, col.regions = "orange") + mapView(
  NY_PM25_point,
  cex = "meanPM25",
  col.regions = "cadetblue1",
  layer.name = "PM2.5 in New York State in 2020",
  map.types = "CartoDB.Positron"
)

#SO2 monitor sites
NY_SO2_Counties_sf = merge(NY_sf_2263, NY_SO2_processed, by.x = "NAME", by.y = "County")
NY_SO2_point = st_centroid(NY_SO2_Counties_sf)
mapView(NY_sf, col.regions = "orange") + mapView(
  NY_SO2_point,
  cex = "meanSO2",
  col.regions = "greenyellow",
  layer.name = "SO2 in New York State in 2020",
  map.types = "CartoDB.Positron"
)

```
